
# R for evolutionary biology: data structures

R is a very flexible programming language, and it allows developers to
create their own data structures (called **classes**) for their packages.
Over the years, some packages have become so popular that the classes
they use to store data are now used the "standard" representations for
particular types of data. Here, we will show some common data classes in
evolutionary biology and comparative genomics research.

## The Bioconductor ecosystem

[The Bioconductor project](https://bioconductor.org/) was created to develop,
support, and disseminate free open source software to analyze biological data.
It works as a repository of R packages for biological data analysis, with
both "core packages" (developed and maintained by the Bioconductor Core Team),
and community-contributed packages. We usually refer to Bioconductor as a
"package ecosystem" because its packages are tightly integrated and designed
for easy interoperation, such that different packages can be used together
with minimal effort. An important facilitator of such interoperability is
the existence of standardized data structures, such as `GRanges` objects
for genomic coordinates, and `DNA/RNA/AAStringSet` objects for sequences.

### The {Biostrings} package and `XStringSet` objects

The Biostrings package [@biostrings] allows efficient manipulation of 
biological sequences, and its data classes are standard for all Bioconductor
packages that involve sequence analyses. The data classes in Biostrings to
represent sets of biological sequences are:

1. `DNAStringSet`: for sets of DNA strings.
2. `RNAStringSet`: for sets of RNA strings.
3. `AAStringSet`: for sets of amino acid strings.

For a single sequence, the each of the above has a correspoding `XString` class
(i.e., `DNAString`, `RNAString`, `AAString`). For example, let's create some
`XString` objects manually:

```{r}
library(Biostrings)

# Create `DNAString` object
dna_seq <- DNAString("ATGGCCGACTCA")
dna_seq

# Convert `DNAString` to `RNAString`
rna_seq <- RNAString(dna_seq)
rna_seq

# Translate `RNAString` to create `AAString`
aa_seq <- translate(rna_seq)
aa_seq
```

In real-world data analyses, we would work with multiple sequences (e.g., 
a whole genome or a whole proteome). Multiple sequences are stored 
in `XStringSet` objects, which can be read from FASTA files 
with `readXStringSet()` functions, namely:

1. `readDNAStringSet()`: read FASTA file containing multiple DNA sequences.
2. `readRNAStringSet()`: read FASTA file containing multiple RNA sequences. 
3. `readAAStringSet()`: read FASTA file containing multiple AA sequences.

For example, let's read an example FASTA file that come with the Biostrings
package.

```{r}
# Path to FASTA file containing an ORF
fasta_file <- system.file("extdata", "someORF.fa", package = "Biostrings")
fasta_file

# Read FASTA file as a DNAStringSet object
dna_seqs <- readDNAStringSet(fasta_file)
dna_seqs
```

Some examples of what you can do to explore `XStringSet` objects include:

```{r}
# width(): get sequence lengths
width(dna_seqs)

# names(): get sequence names
names(dna_seqs)

# subset DNAStringSet and create a DNAStringSet of length 1
dna_seqs[1]

# subset DNAStringSet and create a DNAString object
dna_seqs[[1]]
```

All functions above would also work for `AAStringSet` objects 
and `RNAStringSet` objects.

### The {Biostrings} package and `XMultipleAlignment` objects

Like `XStringSet` objects, `XMultipleAlignment` objects also store sequences,
but sequences that were aligned using a Sequence Alignment algorithm.
FASTA files containing aligned sequences can be read with 
`XMultipleAlignment` functions, namely:

1. `readDNAMultipleAlignment()`: read a FASTA file containing aligned DNA
sequences.
2. `readRNAMultipleAlignment()`: read a FASTA file containing aligned RNA
sequences.
3. `readAAMultipleAlignment()`: read a FASTA file containing aligned amino acid
sequences.


For example:

```{r}
# Path to FASTA file containing MSA of DNA sequences
aln_path <- system.file("extdata", "msx2_mRNA.aln", package = "Biostrings")
aln_path

# Read MSA file
dna_msa <- readDNAMultipleAlignment(aln_path)
dna_msa
```


## The {ape} package and `phylo` objects

The ape package [@ape] is one of the most popular packages for Analyses
of Phylogenetics and Evolution (from where the name comes), and it
represents phylogenies as `phylo` objects. Because of ape's popularity,
`phylo` objects have become the standard data structure for phylogenies in R.

To create `phylo` objects, you will use the function `read.tree()`, which
accepts both the path to a file containing the tree (in Newick format,
for example), or a text representation of a phylogeny. To demonstrate
the structure of a `phylo` object, let's create a phylogeny from text:

```{r}
library(ape)

# Create tree from text
text_tree <- "((((cow, pig),whale), (bat, (lemur, human))), iguana);"
tree <- read.tree(text = text_tree)

tree
```

This is what a `phylo` object looks like. In a sense, a `phylo` object is a 
`list`, a standard R data structure, but with some standard elements. Let's
explore them:

```{r}
# Structure of the `phylo` object
str(tree)

# Exploring each element individually
tree$edge
tree$Nnode
tree$tip.label
```

As `phylo` objects have become standard, there are hundreds of functions
(from dozens of packages) that were especially designed to work on
`phylo` objects. For example, if you want to plot the tree, you would use
the `plot()` function:

```{r}
# Plot tree
plot(tree)
```

We will explore other examples in future lessons.

